Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-26 analysis/paper/paper.Rmd
up to # Materials and archaeological background

Geometric morphometric methods (GMM) has been increasingly applied in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approaches that capture shapes by length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values of coordinate axes. According to different focuses and nature of original shape data, there are two common morphometric methods, landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions, or x, y, z coordinates for three dimensions. Based on distances between raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991]. GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistics analyses [@Slice2007]. Landmark-based morphometrics are widely applied to archaeological objects with a obvious morphological signature that can provide ideal reference points for landmark placement, such as projectile point tips or biologically measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].

However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks are crucial for variation. Furthermore, representative and reliable landmark points may be unavailable for complex or rounded shaped structures. In those cases, outline approaches are better way for analyzing the overall shape of an object. One of the outline approaches is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010]. Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013]. Another approach is elliptic Fourier Analysis (EFA) that turns coordinates into coefficients using mathematical function, commonly applied to two-dimensional closed shape [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adams2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometry of an object. The harmonic power, a cumulated sum of squared harmonic coefficient, provides a robust rule for the desired number of harmonics [@Bonhomme2014].

Outline approaches especially EFA is suitable for shapes lacking representative landmarks or curves convey more meaningful variations. The applications in archaeology are seen on human remains or zooarchaeology to understand the biological variation and recently more on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization. Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico. Their results suggest that some ceramic types from Medio period (AD 1200-1450), were made by specialists based on high standardized shape. Among those high standardized jars, some were made by attached specialists, while others were made by independent specialists according to their spatial distribution patterns. Similarity, @Selden2019 examines 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana from the Clarence H. Webb collections. The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner. Also, two discrete base and body shapes represented north and south difference were identified. For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, the EFA results help to understand the level of production standardization over time across the region. Those examples show that outline approaches can distinguish variation in ceramic shapes in a finer resolution to understand deeply the ceramic production.

Taking the ceramics data from Kiwulan, northeaster Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century that might hint the emergence of ceramic specialization. In addition, we use significance test for the equality of coefficient of variation of shape variables to compare the vessel standarization from different periods in a statistical way. Using pottery shape as a proxy to study craft specialization, we answer the questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels? How does pottery change in shape in relation to foreign influences? Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants? Starting from the historical background of colonial influences and its reflection in the material record, this study explore the relationship between the emergence of social inequality and craft specialization that can extend our understanding of the reactions of indigenous societies in the periphery of colonial control, providing a perspective that could be applied in other parts of the world in a similar context.

>>>>

Geometric morphometric methods (GMM) has been increasingly applied in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interest. Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as xy coordinates for two dimensions, or xyz coordinates for three dimensions. In a typical analysis, starting with raw landmarks, a sample of objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991]. GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistical analyses [@Slice2007]. Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or biological measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].

However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks contains crucial variation. Furthermore, landmark points may be unavailable for structures that are mostly or entirely curves. In those cases, outline approaches are more suitable for analyzing the overall shape of an object. One outline approache is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010]. Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013]. Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].

EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation. Applications in archaeology include human remains and zooarchaeology to understand  biological variation, and more recently on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization. Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico. Their results suggest that some ceramic types from the Medio period (AD 1200-1450), were made by specialists based on highly standardized shapes. Among those highly standardized jars, some were made by attached specialists <!-- what does this mean, 'attached' to what? please clarify -->, while others were made by independent specialists according to their spatial distribution patterns. Similarity, @Selden2019 examined 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana <!-- using semi-landmarks or something else? please add -->. The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner. Also, two discrete base and body shapes represented northern and southern differences were identified. For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek201's EFA results help us understand the level of production standardization over time across the region. Those examples show that outline approaches can distinguish variation in ceramic shapes at high resolutions to understand ceramic production.

Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization. We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standarization from different periods. Using pottery shape as a proxy to study craft specialization, we address these questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels? Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?
